---
title: "Labor Market Analysis EDA"
author: "Rylee Iacolucci, Sophie McDevitt, Caitlin Howansky"
date: "January 20, 2019"
output: html_document
---

#### __Overview__

For our labor marketing analysis we've chosen to analyze Texas. Below are the research questions for this project:

* In Texas, how does earnings vary by household language?
* Does the household language earnings vary by household family type?

To answer these questions we'll be analyzing a person and household data file. The person data file contains over 200,000 observations and 280 columns. To answer our research questions we were able to reduce the data file to ten columns. In the rest of this paper we'll examine this data and describe the filtering out and why it's necessary for our economic analysis.

```{r, include=FALSE, warning=FALSE}
# Name: Rylee Iacolucci
# Date: 1/9/2018
# 
# Description: This is the exploratory data analysis (EDA) for the LMA ECON 5300 Project.
# For this project we've chosen to analyze .... and how it impacts ......
# ...
# ...
# ...

library(tidyverse)
library(here)
library(gridExtra)
library(knitr)
library(kableExtra)
library(formattable)

# tx_pop_df <- read_csv(here("Data/ss16ptx.csv"))
tx_pop_df <- read_csv(here("Data/Person_Record_Adj.csv"))
tx_hh_df <- read_csv(here("Data/HH_Record_Adj.csv"))

# Here I'm going to rename some columns to make doing this eda a little easier.
# For example, PERNP really stands for Total Persons Earnings.
tx_pop_df <- tx_pop_df %>% 
  dplyr::rename(RECORD_TYPE = RT,
                HOUSING_UNIT = SERIALNO,
                PERSON_NUMBER = SPORDER,
                AREA_CODE = PUMA,
                CITIZEN_STATUS = CIT,
                WORKER_CLASS = COW,
                EMPLOYMENT_STATUS = ESR,
                PERSON_EARNINGS = PERNP)

tx_hh_df <- tx_hh_df %>% 
  dplyr::rename(RECORD_TYPE = RT,
                HOUSING_UNIT = SERIALNO,
                AREA_CODE = PUMA,
                HOUSE_LANGUAGE = HHL,
                FAMILY_TYPE = HHT)

# This will be used to make vizualization standard
basic_theme <- 
  theme(legend.position = "none") +
  theme(plot.title = element_text(size=16, face="bold", hjust=0, color="#666666")) +
  theme(axis.title = element_text(size=12, face="bold", color="#666666")) +
  theme(axis.title.y = element_text(angle=90)) +
  theme(axis.title.x = element_text(hjust = .5))


```

```{r, echo=FALSE}
summary(tx_pop_df)

head(tx_pop_df, 10)


```

#### __Summary Statistics__

* AGEP
    +	From the summary data you’re able to see the age of people surveyed has a very wide range, while no nulls appear to be in the dataset.
    + In our analysis we want individuals who are legally able to be fully employed. In Texas the minimum age for full employment is 14, however, by law that individual is still considered a child.
        + People under 18 years old have regulations limiting the total number of hours they can work in a single week.
        + In this analysis we will only analyze people who are not bound by age regulations from the state, so only people greater than 18 years old will only be included.
* Employment Status
    + In this category 1 stands for “Civilian Employed, at Work”, while other codes suggest the person is currently not working full time. In this analysis we may want to filter for only people who are currently employed and working so the results of the analysis are not skewed by invididuals not working and/or not looking for work.
* Person Earnings
    + Right away we noticed there are people with zero or less earnings.
    + According to the Bureau of Labor Statistics nearly 6.2 million workers in Texas were paid the federal minimum wage of $7.25 per hour.
        + This would make a person’s annual earnings of $15,080, without adjusting for federal and local taxes, working 40 hours a week for 52 weeks a year.
        + In this analysis we will only include people who meet the annual earnings of the federal minimum wage without adjusting for taxes.
    + Looking at the first 10 rows of data you see a lot of rounded earnings values like 60,000.
        + In this analysis we may have to keep in mind that when individuals were filling out their income values that they may have rounded their earnings.

    
#### __Additional Summary Data__

```{r, echo=FALSE, fig.width=12, fig.height=5}
tx_pop_df <- tx_pop_df %>% 
  dplyr::filter(EMPLOYMENT_STATUS == 1 & PERSON_EARNINGS > 15080 & AGEP >= 18) %>% 
  dplyr::inner_join(tx_hh_df, c("HOUSING_UNIT", "AREA_CODE"))

t1 <- tx_pop_df %>% 
  dplyr::mutate(HOUSE_LANGUAGE = if_else(HOUSE_LANGUAGE == 1, "English Only",
                                 if_else(HOUSE_LANGUAGE == 2, "Spanish",
                                 if_else(HOUSE_LANGUAGE == 3, "Other Indo-European",
                                 if_else(HOUSE_LANGUAGE == 4, "Asian and Pacific Island",
                                 if_else(HOUSE_LANGUAGE == 5, "Other", "Unknown")))))) %>% 
  dplyr::mutate(MALE = if_else(SEX == 1, 1, 0),
                FEMALE = if_else(SEX == 2, 1, 0)) %>% 
  dplyr::select(HOUSE_LANGUAGE, MALE, FEMALE, PERSON_EARNINGS) %>% 
  dplyr::group_by(HOUSE_LANGUAGE) %>% 
  dplyr::summarise(MALE = sum(MALE),
                   FEMALE = sum(FEMALE),
                   TOTAL_PEOPLE = MALE + FEMALE,
                   AVG_PERSON_EARNINGS = mean(PERSON_EARNINGS)) %>% 
  dplyr::mutate(AVG_PERSON_EARNINGS = color_tile("white", "#3498DB")(AVG_PERSON_EARNINGS)) %>% 
  kable(escape = F, align = "c") %>% 
  kable_styling(bootstrap_options = "striped", "condensed", position = "center", font_size = 13) %>% 
  column_spec(1:5, bold = T, color = "black")



t1


```

##### __Household Language__
Looking at the above table you can see English speaking households are over double the second highest category of Spanish. The average person earnings are comparable through the first four categories but drops significantly for spanish speaking households, while the household with the highest average person earnings being Other Indo-European.

<br>

```{r, echo=FALSE, fig.width=12, fig.height=5}
t2 <- tx_pop_df %>% 
  dplyr::mutate(FAMILY_TYPE = if_else(FAMILY_TYPE == 1, "Married Couple",
                                 if_else(FAMILY_TYPE == 2, "Male No Wife",
                                 if_else(FAMILY_TYPE == 3, "Female No Husband",
                                 if_else(FAMILY_TYPE == 4, "Male Alone",
                                 if_else(FAMILY_TYPE == 5, "Male Not Alone",
                                 if_else(FAMILY_TYPE == 6, "Female Alone",
                                 if_else(FAMILY_TYPE == 7, "Female Not Alone", "Unknown")))))))) %>% 
  dplyr::mutate(MALE = if_else(SEX == 1, 1, 0),
                FEMALE = if_else(SEX == 2, 1, 0)) %>% 
  dplyr::select(FAMILY_TYPE, MALE, FEMALE, PERSON_EARNINGS) %>% 
  dplyr::group_by(FAMILY_TYPE) %>% 
  dplyr::summarise(MALE = sum(MALE),
                   FEMALE = sum(FEMALE),
                   TOTAL_PEOPLE = MALE + FEMALE,
                   AVG_PERSON_EARNINGS = mean(PERSON_EARNINGS)) %>% 
  dplyr::mutate(AVG_PERSON_EARNINGS = color_tile("white", "#3498DB")(AVG_PERSON_EARNINGS)) %>% 
  kable(escape = F, align = "c") %>% 
  kable_styling(bootstrap_options = "striped", "condensed", position = "center", font_size = 13) %>% 
  column_spec(1:5, bold = T, color = "black")

t2

```


##### __Family Type__
The most respondents in this category were Married Couples who also have the highest average earnings compared to other house hold family types. Married couples also make up a significant amount of the respondents. This could yield some interesting result which will non quantifiable variables may have to be considered durning the analysis.

<br>

#### __Graphical Overview__

```{r, echo=FALSE, fig.width=12, fig.height=4.5}
earnings_hist_1 <- tx_pop_df %>% 
  ggplot(aes(x = PERSON_EARNINGS)) +
  geom_histogram(binwidth = 20000, fill = "#880011", color = "white") +
  xlab("Person Earnings") +
  ylab("Frequency") +
  labs(title = "Histogram of Person Earnings",
       subtitle = "Bin Width Every $20,000") +
  basic_theme

earnings_hist_2 <- tx_pop_df %>% 
  ggplot(aes(x = log(PERSON_EARNINGS))) +
  geom_histogram(bins = 20, fill = "#880011", color = "white") +
  xlab("Person Earnings") +
  ylab("Frequency") +
  labs(title = "Histogram of LOG of Person Earnings",
       subtitle = "Grouped Into Twenty Bins") +
  basic_theme


grid.arrange(earnings_hist_1, earnings_hist_2, ncol = 2)



```

#### __Person Earnings__

_Note: Inner joining with the household dataset on HOUSING_UNIT & AREA_CODE removed 151 observations due to NAs._


Looking at the left histogram you can see the person earnings data is right skewed. Some answered their earnings to exceed 500,000, which may be considered an outlier when you compare against the mean of 65,423. In this analysis we don't want to exclude these values since they are not top coded and may provide useful insight in explaining our research questions. Instead we can chose to analyze the log of person earnings. As you can see applying the log function makes the data more normally distributed which may help with analyzing the variance in person earnings.



```{r, echo=FALSE, fig.width=12, fig.height=4.5}
# Here I convert the numerical answers for household languge into it's meanings.
# This is just for the graph, which is why I don't replace the dataframe.
hh_one <- tx_pop_df %>% 
  dplyr::mutate(HOUSE_LANGUAGE = if_else(HOUSE_LANGUAGE == 1, "English Only",
                                 if_else(HOUSE_LANGUAGE == 2, "Spanish",
                                 if_else(HOUSE_LANGUAGE == 3, "Other Indo-European",
                                 if_else(HOUSE_LANGUAGE == 4, "Asian and Pacific Island",
                                 if_else(HOUSE_LANGUAGE == 5, "Other", "Unknown")))))) %>% 
  dplyr::select(HOUSE_LANGUAGE) %>% 
  dplyr::group_by(HOUSE_LANGUAGE) %>% 
  dplyr::summarise(RECORDS = n()) %>% 
  ggplot(aes(x = reorder(HOUSE_LANGUAGE, RECORDS), y = RECORDS)) +
  geom_bar(fill = "#880011", color = "white", stat = "identity") +
  ylab("Frequency") +
  xlab("") +
  labs(title = "Histogram of Household Language") +
  basic_theme +
  coord_flip()


hh_two <- tx_pop_df %>% 
  dplyr::mutate(FAMILY_TYPE = if_else(FAMILY_TYPE == 1, "Married Couple",
                                 if_else(FAMILY_TYPE == 2, "Male No Wife",
                                 if_else(FAMILY_TYPE == 3, "Female No Husband",
                                 if_else(FAMILY_TYPE == 4, "Male Alone",
                                 if_else(FAMILY_TYPE == 5, "Male Not Alone",
                                 if_else(FAMILY_TYPE == 6, "Female Alone",
                                 if_else(FAMILY_TYPE == 7, "Female Not Alone", "Unknown")))))))) %>%
  dplyr::select(FAMILY_TYPE) %>% 
  dplyr::group_by(FAMILY_TYPE) %>% 
  dplyr::summarise(RECORDS = n()) %>% 
  ggplot(aes(x = reorder(FAMILY_TYPE, RECORDS), y = RECORDS)) +
  geom_bar(fill = "#880011", color = "white", stat = "identity") +
  ylab("Frequency") +
  xlab("") +
  labs(title = "Histogram of Family Type") +
  basic_theme +
  coord_flip()

grid.arrange(hh_one, hh_two, ncol = 2)





```

#### __Family Type__

Both the histgrams above show the answer frequency for the household language and family type of the person surveyed. You can see english and married couple households are the majority of those surveyed. The impact of household and family types will be examined in our technical research paper.



























