---
title: "LMA Analysis"
author: "Rylee Iacolucci"
date: "January 30, 2019"
output: 
  html_document:
    toc : true
---


```{r, include = FALSE, warning = FALSE}
# This is our first analysis of the LMA data. In this code we'll be
# performing different statistical analyze, which is different from
# the EDA code.
library(tidyverse)
library(here)
library(gridExtra)
library(knitr)
library(kableExtra)
library(formattable)
library(scales)
library(lmtest)
library(sandwich)

tx_pop_df <- read.csv(here("Data/tx_pop_data_v2.csv"))

# This will be used to make vizualization standard
basic_theme <- 
  theme(legend.position = "none") +
  theme(plot.title = element_text(size=16, face="bold", hjust=0, color="#000000")) +
  theme(axis.title = element_text(size=12, face="bold", color="#000000")) +
  theme(axis.title.y = element_text(angle=90)) +
  theme(axis.title.x = element_text(hjust = .5)) +
  theme_minimal() +
  theme(axis.line = element_line(color = "#000000"),
        axis.ticks = element_line(color = "#000000"),
        axis.text = element_text(color = "#000000"))

# Filter For Working Age Persons
# $418,000 is the top code value for Texas
tx_pop_df <- tx_pop_df %>% 
  dplyr::filter(EMPLOYMENT_STATUS == 1 & WAGE > 15080 & WAGE < 418000 & AGEP >= 18 & AGEP < 62)

# Original data source column names:
# 
# person_number: SPORDER
# citizen_status: CIT
# class_of_worker: COW
# employment_status: ESR
# person_earnings: PERNP
# educational_attainment: SCHL
# marital_status: MAR
# times_married: MARHT

tx_pop_df <- 
  tx_pop_df %>% 
  mutate(SEX = if_else(SEX == 1, "male", "female"),
         MARITAL_STATUS = if_else(MARITAL_STATUS == 1, "married",
                          if_else(MARITAL_STATUS == 2, "widowed",
                          if_else(MARITAL_STATUS == 3, "divorced",
                          if_else(MARITAL_STATUS == 4, "separated",
                          if_else(MARITAL_STATUS == 5, "never Married", "unknown"))))),
         EDUCATIONAL_ATTAINMENT = if_else(EDUCATIONAL_ATTAINMENT == 1, "no_schooling",
                                  if_else(EDUCATIONAL_ATTAINMENT < 16, "primary_education",
                                  if_else(EDUCATIONAL_ATTAINMENT < 18, "ged_or_equivalent",
                                  if_else(EDUCATIONAL_ATTAINMENT < 20, "some_college",
                                  if_else(EDUCATIONAL_ATTAINMENT == 20, "associates",
                                  if_else(EDUCATIONAL_ATTAINMENT == 21, "bachelors",
                                  if_else(EDUCATIONAL_ATTAINMENT == 22, "masters",
                                  if_else(EDUCATIONAL_ATTAINMENT == 23, "professional",
                                  if_else(EDUCATIONAL_ATTAINMENT == 24, "doctorate", "Other"))))))))))

# This takes the long dataset and makes it wide by taking the education 
# columns and making them into their own column. I will do the same for
# sex.
tx_pop_df_spread <- 
  tx_pop_df %>% 
  mutate(value = 1) %>% 
  spread(EDUCATIONAL_ATTAINMENT, value, fill = 0)

tx_pop_df_spread <- 
  tx_pop_df_spread %>% 
  mutate(value = 1) %>% 
  spread(SEX, value, fill = 0)


```

## __Data Overview__

```{r, echo=FALSE, fig.width=12, fig.height=5}
g1 <- 
  tx_pop_df %>% 
  ggplot(aes(x = WAGE)) +
  geom_histogram(bins = 20, fill = "#880011", color = "white") +
  basic_theme +
  ggtitle("Wage Histogram",
          sub = "Normal Distribution") +
  scale_y_continuous(label = comma, limits = c(0, 20000))

g2 <- 
  tx_pop_df %>% 
  ggplot(aes(x = log(WAGE))) +
  geom_histogram(bins = 20, fill = "#880011", color = "white") +
  basic_theme +
  ggtitle("Log Wage Histogram",
          sub = "More Normal Distribution") +
  scale_y_continuous(label = comma, limits = c(0, 10000)) +
  xlab("Log Of Wage")

grid.arrange(g1, g2, ncol = 2)

```

## __BPG Test__

```{r}

model_one <- lm(WAGE ~ primary_education + ged_or_equivalent + some_college + associates + bachelors + masters + professional + doctorate + AGEP + male, data = tx_pop_df_spread)

bptesteq = lm(residuals(model_one)*residuals(model_one) ~ primary_education + ged_or_equivalent + some_college + associates + bachelors + masters + professional + doctorate + AGEP + male, data = tx_pop_df_spread)
summary(bptesteq)



```
Since the p-value is so small, we reject the null hypothesis that the variances are equal across all y. Therefore, we have stong evidence of heteroskedasticity. To correct for this, we will take the log of wage.


## __Model One__

```{r, echo=FALSE}
model_one_log <- lm(log(WAGE, base = exp(1)) ~ primary_education + ged_or_equivalent + some_college + associates + bachelors + masters + professional + doctorate + AGEP + I(AGEP*AGEP)+ male, data = tx_pop_df_spread)



bptesteq = lm(residuals(model_one_log)*residuals(model_one) ~ primary_education + ged_or_equivalent + some_college + associates + bachelors + masters + professional + doctorate + (AGEP*AGEP) + male, data = tx_pop_df_spread)
summary(bptesteq)

```
In this model, we still have heteroskedasticity so we will have to take the robust standard errors. 


```{r}
#Generate robust standard errors

vcovHC(model_one_log, type = "HC")
modone_log_se <- diag(vcovHC(model_one_log, type = "HC"))^0.5
modone_log_se

```


Holding all other variables constant someone with a bachelors degree on average earns `r round((exp(0.7337389) - 1) * 100, 2)`% more than someone with no education. Someone with a professional degree earns on average `r round((exp(1.0730300) - 1) * 100, 2)`% more than someone with no education.


```{r, echo=FALSE}
t1 <- 
  tx_pop_df %>% 
  select(EDUCATIONAL_ATTAINMENT, SEX, AGEP, WAGE) %>% 
  group_by(EDUCATIONAL_ATTAINMENT, SEX) %>% 
  summarise(`Avg Age` = mean(AGEP),
            `Avg Wage` = mean(WAGE)) %>% 
  mutate(`Avg Wage` = dollar(`Avg Wage`)) %>%
  kable(size = 10, align = "l", escape = F) %>% 
  kable_styling(bootstrap_options = c("striped", "condensed")) %>% 
  column_spec(1:4, color = "#000000") %>% 
  column_spec(1, width = "12em") %>% 
  collapse_rows(1, valign = "top")
  

t1


```



















