---
title: "BPG Tests and Linear Models"
author: "Caitlin Howansky"
date: "February 4, 2019"
output:
  word_document:
    toc: yes
  html_document:
    toc: yes
---

For our labor marketing analysis we've chosen to analyze Texas. Below are the research questions for this project:

* In Texas how do earnings vary by educational attainment?
* How does the premium for education vary by marital status?

To answer these questions we'll be analyzing the personal data file. The person data file contains over 200,000 observations and 280 columns. To answer our research questions we were able to reduce the data file to ten columns. In the rest of this paper we'll examine this data and describe the filtering out and why it's necessary for our economic analysis.

In our initial EDA we found wide range of values for person earnings. Examining basic Texas work regulations we found the minimum age for full employment was 14 year old, however, people below 18 have restrictions on the amount of hours they're allowed to work in a single week. In our analysis we want to examine people who are employed and not resticted by state regulations, which means we'll only be examining people fully employed and at least 18 years old. We are also filter for people that meet the annual earnings if they were receiving the federal minimum wage, which is $15,080 without adjusting for federal, state, and local taxes.



```{r, include = FALSE, warning = FALSE}
library(tidyverse)
library(here)
library(gridExtra)
library(knitr)
library(kableExtra)
library(formattable)
library(scales)
library(lmtest)
library(sandwich)

tx_pop_df <- read.csv(here("Data/tx_pop_data_v2.csv"))

# This will be used to make vizualization standard
basic_theme <- 
  theme(legend.position = "none") +
  theme(plot.title = element_text(size=16, face="bold", hjust=0, color="#000000")) +
  theme(axis.title = element_text(size=12, face="bold", color="#000000")) +
  theme(axis.title.y = element_text(angle=90)) +
  theme(axis.title.x = element_text(hjust = .5)) +
  theme_minimal() +
  theme(axis.line = element_line(color = "#000000"),
        axis.ticks = element_line(color = "#000000"),
        axis.text = element_text(color = "#000000"))

# Filter For Working Age Persons
# $418,000 is the top code value for Texas
tx_pop_df <- tx_pop_df %>% 
  dplyr::filter(EMPLOYMENT_STATUS == 1 & WAGE > 15080 & WAGE < 418000 & AGEP >= 18 & AGEP < 62)

# Original data source column names:
# 
# person_number: SPORDER
# citizen_status: CIT
# class_of_worker: COW
# employment_status: ESR
# person_earnings: PERNP
# educational_attainment: SCHL
# marital_status: MAR
# times_married: MARHT

tx_pop_df <- 
  tx_pop_df %>% 
  mutate(LOG_WAGE = log(WAGE),
         SEX = if_else(SEX == 1, "male", "female"),
         MARITAL_STATUS = if_else(MARITAL_STATUS == 1, "married",
                          if_else(MARITAL_STATUS == 2, "widowed",
                          if_else(MARITAL_STATUS == 3, "divorced",
                          if_else(MARITAL_STATUS == 4, "separated",
                          if_else(MARITAL_STATUS == 5, "never Married", "unknown"))))),
         EDUCATIONAL_ATTAINMENT = if_else(EDUCATIONAL_ATTAINMENT == 1, "no_schooling",
                                  if_else(EDUCATIONAL_ATTAINMENT < 16, "primary_education",
                                  if_else(EDUCATIONAL_ATTAINMENT < 18, "ged_or_equivalent",
                                  if_else(EDUCATIONAL_ATTAINMENT < 20, "some_college",
                                  if_else(EDUCATIONAL_ATTAINMENT == 20, "associates",
                                  if_else(EDUCATIONAL_ATTAINMENT == 21, "bachelors",
                                  if_else(EDUCATIONAL_ATTAINMENT == 22, "masters",
                                  if_else(EDUCATIONAL_ATTAINMENT == 23, "professional",
                                  if_else(EDUCATIONAL_ATTAINMENT == 24, "doctorate", "Other"))))))))),
         RACE = if_else(RACE == 1, "WHITE",
                if_else(RACE == 2, "AFRICAN_AMERICAN",
                if_else(RACE == 3, "AMERICAN_INDIAN", "OTHER"))))

# This takes the long dataset and makes it wide by taking the education 
# columns and making them into their own column. I will do the same for
# sex.
tx_pop_df_spread <- 
  tx_pop_df %>% 
  mutate(value = 1) %>% 
  spread(EDUCATIONAL_ATTAINMENT, value, fill = 0)

tx_pop_df_spread <- 
  tx_pop_df_spread %>% 
  mutate(value = 1) %>% 
  spread(SEX, value, fill = 0)

tx_pop_df_spread <- 
  tx_pop_df_spread %>% 
  mutate(value = 1) %>% 
  spread(RACE, value, fill = 0)


```


## __BPG Test__

For our model, we have chosen to use educational attainment, age, sex, and race as the independent variables describing wage. Later we'll examine whether including age, sex, and race are enough variables to satisfy our model not violating the zero conditional mean assumption. Since we are examine wage and know that our data is right skewed we suspect our model to have heteroskedasticity. To test whether or not this is true we'll use a BPG test.

### __Wage Model for Males__

```{r, echo = TRUE}

# This just filters for all males in the dataset. We're chosing to seperate females and males
# to make the interpretation of our results easier.
tx_pop_df_spread_male <- 
  tx_pop_df_spread %>% 
  filter(male == 1)

# A person with no education is left out.
# A person who identifies their race as other is left out.
model_one_male <- lm(WAGE ~ (primary_education + ged_or_equivalent + some_college + associates + bachelors + 
                        masters + professional + doctorate + AGEP + AFRICAN_AMERICAN + AMERICAN_INDIAN + WHITE), data = tx_pop_df_spread_male)

bptest(model_one_male)

bptesteq = lm(residuals(model_one_male) * residuals(model_one_male) ~ 
                primary_education + ged_or_equivalent + some_college + associates + bachelors + masters + professional + doctorate + AGEP, data = tx_pop_df_spread_male)

# This is the summary of the first model. Without taking the LOG of wage.
# We know this will not yield the best results.
# summary(model_one)

```



```{r, echo=FALSE}

# BPG test for first model.
summary(bptesteq)

```



Since the p-value is so small, we reject the null hypothesis that the variances are equal across all y. Therefore, we have stong evidence of heteroskedasticity. To correct for this, we will take the log of wage.




```{r, echo=TRUE}
model_one_log <- lm(LOG_WAGE ~ (primary_education + ged_or_equivalent + some_college + associates + bachelors + 
                                masters + professional + doctorate + AGEP + AGEP ^ 2 + AFRICAN_AMERICAN + AMERICAN_INDIAN + WHITE), data = tx_pop_df_spread_male)

bptesteq = lm(residuals(model_one_log) * residuals(model_one_log) ~ 
              (primary_education + ged_or_equivalent + some_college + associates + bachelors + 
               masters + professional + doctorate + AGEP ^ 2 + AFRICAN_AMERICAN + AMERICAN_INDIAN + WHITE), data = tx_pop_df_spread_male)

#summary(bptesteq)

bptest(model_one_log)
```

In this model, since the P value is still close to zero we have strong evidence of heteroskedasticity so we will have to take the robust standard errors. 


```{r, echo = TRUE}

#Generate robust standard errors
vcovHC(model_one_log, type = "HC")

modone_log_se <- diag(vcovHC(model_one_log, type = "HC")) ^ 0.5

modone_log_se

```

### __Wage Model for Females__

```{r, echo = TRUE}
tx_pop_df_spread_female <- 
  tx_pop_df_spread %>% 
  filter(female == 1)

model_one_female <- lm(WAGE ~ (primary_education + ged_or_equivalent + some_college + associates + bachelors + 
                        masters + professional + doctorate + AGEP + AFRICAN_AMERICAN + AMERICAN_INDIAN + WHITE), data = tx_pop_df_spread_female)

bptesteq = lm(residuals(model_one_female)*residuals(model_one_female) ~ 
             (primary_education + ged_or_equivalent + some_college + associates + bachelors + 
              masters + professional + doctorate + AGEP + AFRICAN_AMERICAN + AMERICAN_INDIAN + WHITE), data = tx_pop_df_spread_female)

#summary(bptesteq)

bptest(model_one_female)
```

Again, because the p-value is so small, we reject the null hypothesis so that we have strong evidence to say that this model is heteroskedastic. 



```{r, echo=TRUE}

model_two_log <- lm(LOG_WAGE ~ (primary_education + ged_or_equivalent + some_college + associates + bachelors + 
                                masters + professional + doctorate + AGEP + AGEP ^ 2 + AFRICAN_AMERICAN + AMERICAN_INDIAN + WHITE), data = tx_pop_df_spread_female)

bptesteq = lm(residuals(model_two_log) * residuals(model_two_log) ~ 
             (primary_education + ged_or_equivalent + some_college + associates + bachelors + 
              masters + professional + doctorate + AGEP ^ 2 + AFRICAN_AMERICAN + AMERICAN_INDIAN + WHITE), data = tx_pop_df_spread_female)

#summary(bptesteq)
bptest(model_two_log)
```

Since there is still heteroskedasticity, we will correct with the robust standard errors. 


```{r, echo = TRUE}
#Generate robust standard errors

vcovHC(model_two_log, type = "HC")
modone_log_se <- diag(vcovHC(model_two_log, type = "HC")) ^ 0.5
modone_log_se


```

Our next models include marital status.


```{r, echo = TRUE}
tx_pop_df_spread <- 
  tx_pop_df_spread %>% 
  mutate(value = 1) %>% 
  spread(MARITAL_STATUS, value, fill = 0)

tx_pop_df_spread_male <- 
  tx_pop_df_spread %>% 
  filter(male == 1)

# Never married left out of model.
model_married_male <- lm(WAGE ~ (primary_education + ged_or_equivalent + some_college + associates + bachelors + 
                                 masters + professional + doctorate + AGEP + AFRICAN_AMERICAN + AMERICAN_INDIAN + WHITE +
                                 divorced + married + separated + widowed), data = tx_pop_df_spread_male)

bptesteq = lm(residuals(model_married_male) * residuals(model_married_male) ~ 
             (primary_education + ged_or_equivalent + some_college + associates + bachelors + 
              masters + professional + doctorate + AGEP + AFRICAN_AMERICAN + AMERICAN_INDIAN + WHITE +
              divorced + married + separated + widowed), data = tx_pop_df_spread_male)

#summary(bptesteq)
bptest(model_married_male)
```

Because of the heteroskedasticity, we will take the log of the wage variable. 


```{r, echo = TRUE}
model_male_log <- lm(LOG_WAGE ~ (primary_education + ged_or_equivalent + some_college + associates + bachelors + 
                                 masters + professional + doctorate + AGEP + AGEP ^ 2 + AFRICAN_AMERICAN + AMERICAN_INDIAN + WHITE +
                                 divorced + married + separated + widowed), data = tx_pop_df_spread_male)


bptesteq = lm(residuals(model_male_log) * residuals(model_male_log) ~ 
             (primary_education + ged_or_equivalent + some_college + associates + bachelors + 
              masters + professional + doctorate + AGEP ^ 2 + AFRICAN_AMERICAN + AMERICAN_INDIAN + WHITE +
              divorced + married + separated + widowed), data = tx_pop_df_spread_male)

#summary(bptesteq)
bptest(model_male_log)
```

Generating robust standard errors:


```{r, echo = TRUE}
vcovHC(model_male_log, type = "HC")
modone_log_se <- diag(vcovHC(model_male_log, type = "HC"))^0.5
modone_log_se
```

